Randomized trace estimation of parameter-dependent matrices applied to spectral density approximation

### 2024-05-24 paper/main.tex
add stable rank discussion

In numerous problems across physics, chemistry, engineering, and data science, eigenvalues and especially their distribution are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be captured with the spectral density
\begin{equation}
    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i).
\end{equation}
Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. In many applications, it is also acceptable to trade in some of the finer details of the spectrum for a better approximability by smoothing $\phi$, typically through convolution with a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This converts the problem to estimating the trace of a smooth parameter-dependent matrix function.

\todo{
\begin{itemize}
    \item Definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly \cite{pearlmutter-1994-fast-exact} or for complicated matrix functions [partition function, spectral density, exponential integrators])
    \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
    \item Trace estimation for parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation but only discrete setting [Woodruff, Dharangutte])
    %\item Error analysis for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
\end{itemize}

\paragraph{Contribution}
\begin{itemize}
    \item Theoretical guarantees in the $L^1$-norm for three trace estimators on parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
    \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators using Chebyshev expansion, including theoretical guarantees
\end{itemize}
}

\paragraph{Notation}
We follow the notation and conventions which are used in contemporary literature in this field. Further,
\begin{itemize}
    \item we denote with $\mathbb{N}$ all positive whole numbers and with $\mathbb{R}$ all real numbers;
    \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
    \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots),
    vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots),
    and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
    \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$. The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$;
    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$
    carries ones on its diagonal and zeros everywhere else. The zero matrix
    $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
    \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
    \item matrix norms are denoted with $\lVert \cdot \rVert$. For symmetric matrices, the $p$-Schatten norm is
    $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$. We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
\end{itemize}

@@

\paragraph{$L^{1}$ norm}
We define the $L^{1}$ norm of a square-integrable function $f:[a,b] \to \mathbb{R}$ as
\begin{equation}
    \int_{a}^{b} | f(t) | \mathrm{d}t.
\end{equation}
\todo{Motivate why we use this.}

@@

We first consider \refequ{equ:hutchinson-trace-estimator}.
\todo{Discuss theory for constant matrix
}

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query quadratic trace estimate 
    \begin{equation}
        r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Trace(\mtx{B}(t))
    \end{equation}
    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

    First, we consider $r(t)$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$. Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments of $r$ for a fixed $t$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 16 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\sfrac{1}{2 k}}~\mathrm{d}t
        < 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    By Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(t)|~\mathrm{d}t < 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\mtx{\Psi}}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \lVert \mtx{A}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ and
    \begin{equation}
        r(t) = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
    \end{equation}
    Hence, we conclude from \refequ{equ:hutchinson-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
        < 16 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}

@@

\todo{Discuss $1/\varepsilon^2$ drawback}

@@

Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank}

Pointwise eigenvalue decomposition
\begin{equation}
    \mtx{A}(t) 
    = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
\end{equation}
where $\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$.

We define the matrix partitions
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \quad \text{and} \quad
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.

@@

We consider the task of computing the spectral density of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, which is defined as
\begin{equation}
    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i),
    \label{equ:spectral-density}
\end{equation}
where $\delta$ is the Dirac delta distribution.

\subsection{Introduction to spectral density estimation}
\label{subsec:spectral-density}

>>>>

In numerous problems across physics, chemistry, engineering, and data science, eigenvalues and especially their distribution are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be captured in the spectral density
\begin{equation}
    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i).
\end{equation}
Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. In many applications, it is also acceptable to trade in some of the finer details of the spectrum for a better approximability by smoothing $\phi$, typically through convolution with a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This converts the problem to estimating the trace of a smooth parameter-dependent matrix.

\todo{
\begin{itemize}
    \item Brief definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly \cite{pearlmutter-1994-fast-exact} or for complicated matrix functions [partition function, spectral density, exponential integrators])
    \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator} and variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
    \item Trace estimation in the parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, only low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation for sequence of similar matrices \cite{woodruff-2022-optimal-query, dharangutte-2021-dynamic-trace})
    %\item Error analysis for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
\end{itemize}

\paragraph{Contribution}
\begin{itemize}
    \item Theoretical guarantees in the $L^1$-norm for three established trace estimators for constant matrices when applied to parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
    \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators using Chebyshev expansion, including theoretical guarantees
\end{itemize}
}

\paragraph{Notation}
We follow the notation and conventions which are used in contemporary literature in this field. Further,
\begin{itemize}
    \item we denote with $\mathbb{N}$ all positive whole numbers;
    \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
    \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots), vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots), and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
    \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$. The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$;
    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$ carries ones on its diagonal and zeros everywhere else. The zero matrix $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
    \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
    \item matrix norms are denoted with $\lVert \cdot \rVert$. For symmetric matrices, the $p$-Schatten norm is $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$. We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
\end{itemize}

@@

\paragraph{$L^{1}$ norm}
We define the $L^{1}$ norm of a square-integrable function $f:[a,b] \to \mathbb{R}$ as
\begin{equation}
    \int_{a}^{b} | f(t) | \mathrm{d}t.
\end{equation}
We use it to measure the error of the trace estimators. \todo{Motivate why we use this.}

@@

We first consider the estimator \refequ{equ:hutchinson-trace-estimator}.

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 2^4 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query quadratic trace estimate 
    \begin{equation}
        r(t) = \Trace(\mtx{B}(t)) - \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi}
    \end{equation}
    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

    First, we consider $r(t)$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$. Thus, by \cite[theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments of $r$ for a fixed $t$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 2^4 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Minkowski's integral inequality \cite[theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}^{2 k}\left[ r(t) \right]~\mathrm{d}t
        < 2^4 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    By Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(t)|~\mathrm{d}t < 2^4 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\mtx{\Psi}}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \lVert \mtx{A}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ and
    \begin{equation}
        r(t) = \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
    \end{equation}
    Hence, we conclude from \refequ{equ:hutchinson-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j \right| ~ \mathrm{d}t
        < 2^4 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}

@@

\begin{remark}
    If additionally $\mtx{A}(t)$ is nonzero for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
    \begin{equation}
        \int_{a}^{b} \frac{| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) |}{\Trace(\mtx{A}(t))} ~\mathrm{d}t < \varepsilon,
    \end{equation}
    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{A}(t) \rVert _2 / \Trace(\mtx{A}(t))$. Compared to the constant matrix equivalent \cite[remark 2]{cortinovis-2022-randomized-trace} we observe an additional factor of $\log(\delta^{-1})$.
\end{remark}

\todo{The drawback of the Girard-Hutchinson estimator is the $\mathcal{O}(\varepsilon^{-2})$-requirement on the number of queries. }

@@

Next we consider \refequ{equ:nystrom-trace-estimator}. \todo{Motivate Nyström approximation if matrix is low-rank.}

\begin{proof}
    Since $f$ is non-negative, $f(\mtx{A}, t)$ is positive definite for all $\mtx{A}$ and $t$.
    Notice that by \cite[lemma 2.1]{frangella-2023-randomized-nystrom}, $f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)$ is also positive semi-definite for all $t$. Therefore,
    \begin{equation}
        \left| \Trace(f(\mtx{A}, t)) - \Trace(\widehat{f}(\mtx{A}, t)) \right|
        = \left| \Trace(f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t)) \right|
        = \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast}
    \end{equation}
    From the proof of \cite[corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
    \begin{equation}
        \lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2
    \end{equation}
    Hence, by \cite[theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows
    with $\lVert \cdot \rVert _2 \leq \lVert \cdot \rVert _F$ and $\sigma_i(f^{\sfrac{1}{2}}(\mtx{A}, t))^{2} = \sigma_i(f(\mtx{A}, t))$.
\end{proof}

\todo{Explain that \reflem{lem:nystrom} is a weaker result, and mainly there to help proof \refthm{thm:nystrom-pp}.}

For the coming pages, we employ the following notation. Let
\begin{equation}
    \mtx{A}(t) 
    = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
\end{equation}
be the eigenvalue decomposition of $\mtx{A}(t)$ at a fixed $t$, i.e. $\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$ and $\mtx{U}(t)^{\top} \mtx{U}(t) = \mtx{I}_n$. We define the matrix partitions
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+2ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \quad \text{and} \quad
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.

@@

We consider the task of computing the spectral density of a real symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ with eigenvalues $\lambda_1, \dots, \lambda_n$, which is defined as
\begin{equation}
    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i),
    \label{equ:spectral-density}
\end{equation}
where $\delta$ is the Dirac delta distribution. Finding approximations to this distribution is challenging. Furthermore, in most applications the exact location of every eigenvalue is not important, but more so their approximate locations relative to each other, such as eigenvalue clusters, spectral gaps, or outliers. In the following we will introduce a smooth version of the spectral density, which is significantly easier to approximate at the cost of losing some of the finer characteristics of the spectrum.

\subsection{Smooth spectral density}
\label{subsec:spectral-density}
